---
title: "R Notebook"
output: html_notebook
---

This plot primarily uses the barplot() function in R. You can see the full documentation by entering ?barplot in the console, or visiting https://www.rdocumentation.org/packages/graphics/versions/3.6.2/topics/barplot. For some details on customization, see https://www.r-graph-gallery.com/210-custom-barplot-layout.html.


Read in the data:

```{r}
table4 <- read.table(file = "Table4_Percentage_public_school_districts_by_activity.txt", 
               sep="\t",
               header = TRUE)
```

```{r}
table4
```

Since the percent variable is a decimal, and we want to display it as a percentage in the visualization, the following calculation is performed on the field:

```{r}
table4$pctPublicSchoolDistricts <- table4$pctPublicSchoolDistricts*100
table4
```


Basic bar chart:

```{r}
bar_values <- table4$pctPublicSchoolDistricts
bar_names <- table4$characteristic

barplot(height = bar_values, names.arg = bar_names)
```


To make the plot horizontal: 

```{r}
bar_values <- table4$pctPublicSchoolDistricts
bar_names <- table4$characteristic

barplot(height = bar_values, names.arg = bar_names, 
        horiz = TRUE)
```


Making the plot horizontal reverses the order of the bars. To fix this, reverse the value and name vectors: 

```{r}
bar_values <- rev(table4$pctPublicSchoolDistricts)
bar_names <- rev(table4$characteristic)

barplot(height = bar_values, names.arg = bar_names,
        horiz = TRUE)
```


To make label text horizontal:

```{r}
bar_values <- rev(table4$pctPublicSchoolDistricts)
bar_names <- rev(table4$characteristic)

barplot(height = bar_values, names.arg = bar_names,
        las = TRUE, horiz = TRUE)
```


To adjust plot margins and set figure size for viewer:

```{r, fig.width=20, fig.height=5}
# the par() function is used to set graphical parameters, like margins
# oma is for setting outer margin area space, mar for plot margins
# these set the figure dimensions in inches
# order of vector is bottom, left, top, right

# remove plot margins to fit plot content better, add outer margins to prevent label text from being cut off and leave room for plot title::
par(omi=c(0,0,0,0), mai=c(0.25,8,1.5,0.25))

bar_values <- rev(table4$pctPublicSchoolDistricts)
bar_names <- rev(table4$characteristic)

barplot(height = bar_values, names.arg = bar_names,
        las = TRUE, horiz = TRUE)
```


To change text size and font:

```{r, fig.width=20, fig.height=5}
# adjust left outer margin with font size to prevent text being cut off
par(omi=c(0,0,0,0), mai=c(0.25,12.25,1.5,0.25))

bar_values <- rev(table4$pctPublicSchoolDistricts)
bar_names <- rev(table4$characteristic)

barplot(height = bar_values, names.arg = bar_names,
        las = TRUE, horiz = TRUE,
        cex.names = 1.5, family = "sans")

# for a large font selection, download the "extrafonts" package
```


To change bar width and color:

```{r, fig.width=20, fig.height=5}
par(omi=c(0,0,0,0), mai=c(0.25,12.25,1.5,0.25))

bar_values <- rev(table4$pctPublicSchoolDistricts)
bar_names <- rev(table4$characteristic)
bar_color <- "#7cc354"

barplot(height = bar_values, names.arg = bar_names,
        las = TRUE, horiz = TRUE,
        cex.names = 1.5, family = "sans",
        col = bar_color, border = "transparent",
        space = 0.5)

# "col" can also be set to a vector of colors for different colored bars
```


To format plot style and axes:

```{r, fig.width=20, fig.height=5}
par(omi=c(0,0,0,0), mai=c(0.25,12.25,1.5,0.25))

bar_values <- rev(table4$pctPublicSchoolDistricts)
bar_names <- rev(table4$characteristic)
bar_color <- "#7cc354"


barplot(height = bar_values, names.arg = bar_names,
        las = TRUE, horiz = TRUE,
        cex.names = 1.5, family = "sans",
        col = bar_color, border = "transparent",
        space = 0.5, axes = FALSE)

# full documentation for axis function: https://www.rdocumentation.org/packages/graphics/versions/3.5.2/topics/axis
# 1 for horizontal axis and 2 for vertical axis
# "lwd.tick" = 0 turns off tick marks by setting their line width to 0
# the "at" vector sets the range (length) of the axis line
# "lwd"  sets the axis line width
axis(2, lwd.tick=0, labels=FALSE, at=c(0,100), lwd = 3)
```


To add a title:

```{r, fig.width=20, fig.height=5}
par(omi=c(0,0,0,0), mai=c(0.25,12.25,1.5,0.25))

bar_values <- rev(table4$pctPublicSchoolDistricts)
bar_names <- rev(table4$characteristic)
bar_color <- "#7cc354"
plt_title <- "Among public school districts that offered CTE\nprograms, percentage by activity offered"

barplot(height = bar_values, names.arg = bar_names,
        las = TRUE, horiz = TRUE,
        cex.names = 1.5, family = "sans",
        col = bar_color, border = "transparent",
        space = 0.5, axes = FALSE,
        main = plt_title,
        cex.main = 2)

axis(2, lwd.tick=0, labels=FALSE, at=c(0,100), lwd = 3)
```


For more control over title position, put title in outer margin of plot:

```{r, fig.width=20, fig.height=5}
# adjust margin values to switch extra space for title from plot margin to outer margin
par(omi=c(0,0,1.25,0), mai=c(0.25,12.25,0.25,0.25))

bar_values <- rev(table4$pctPublicSchoolDistricts)
bar_names <- rev(table4$characteristic)
bar_color <- "#7cc354"
plt_title <- "Among public school districts that offered CTE\nprograms, percentage by activity offered"

p <- barplot(height = bar_values, names.arg = bar_names,
        las = TRUE, horiz = TRUE,
        cex.names = 1.5, family = "sans",
        col = bar_color, border = "transparent",
        space = 0.5, axes = FALSE)
        
title(main = plt_title,
        cex.main = 2.25, family = "sans", outer = TRUE, line = 0.75)
# "outer" moves title to outer margin of plot and "line" value controls verical positioning

axis(2, lwd.tick=0, labels=FALSE, at=c(0,100), lwd = 3)
```


To add bar labels:

```{r, fig.width=20, fig.height=5}
par(omi=c(0,0,1.25,0), mai=c(0.25,12.25,0.25,0.25))

bar_values <- rev(table4$pctPublicSchoolDistricts)
bar_names <- rev(table4$characteristic)
bar_color <- "#7cc354"
plt_title <- "Among public school districts that offered CTE\nprograms, percentage by activity offered"

# the paste0() function joins multiple strings together (in this case, a percent sign is joined to each data value)
# the round() function with digits set to 0 rounds each data value to a whole number
text_labels <- paste0(round(bar_values, digits = 0),"%")

p <- barplot(height = bar_values, names.arg = bar_names,
        las = TRUE, horiz = TRUE,
        cex.names = 1.5, family = "sans",
        col = bar_color, border = "transparent",
        space = 0.5, axes = FALSE, 
        xlim=c(0,max(bar_values)+6))
# specifying x-axis range leaves room for bar label text
        
title(main = plt_title,
        cex.main = 2.25, family = "sans", outer = TRUE, line = 0.75)

axis(2, lwd.tick=0, labels=FALSE, at=c(0,100), lwd = 3)

#Set the text labels to be slightly to the right of the actual values by adding 3
text(y = p, x = bar_values + 3, cex = 1.25, 
     labels = text_labels)
```


To save the plot:

```{r}
# use same figure size we have been using
# you must specify inches for png or jpeg (the default is pixels)
png("example_piechart.png", width = 20, height = 5, units="in", res=300)

#Can also save as a jpeg or pdf
# jpeg("example_piechart.jpg", width = 7, height = 5, units="in", res=300)
# pdf("example_barplot.pdf", width=20, height=5)



par(omi=c(0,0,1.25,0), mai=c(0.25,12.25,0.25,0.25))

bar_values <- rev(table4$pctPublicSchoolDistricts)
bar_names <- rev(table4$characteristic)
bar_color <- "#7cc354"
plt_title <- "Among public school districts that offered CTE\nprograms, percentage by activity offered"
text_labels <- paste0(round(bar_values, digits = 0),"%")

p <- barplot(height = bar_values, names.arg = bar_names,
        las = TRUE, horiz = TRUE,
        cex.names = 1.5, family = "sans",
        col = bar_color, border = "transparent",
        space = 0.5, axes = FALSE, 
        xlim=c(0,max(bar_values)+6))
        
title(main = plt_title,
        cex.main = 2.25, family = "sans", outer = TRUE, line = 0.75)

axis(2, lwd.tick=0, labels=FALSE, at=c(0,100), lwd = 3)

text(y = p, x = bar_values + 3, cex = 1.25, 
     labels = text_labels)

dev.off()
```

"null device 1" should appear as output. This indicates that R has finished saving the file, and the plot is no longer available in your environment.
