---
title: "Ggplot2 example"
output: html_document
---

```{r}
library(ggplot2)
library(plotly)
```

Read in data:
```{r}
table1 <- read.table(file = "../Table1_English_language_learner_status_CTE_participants.txt", 
               sep="\t",
               header = TRUE)

```

Convert decimal to percentage and add category column:
```{r}
table1$percentageCTEparticipants <- table1$percentageCTEparticipants*100
table1$category <- c("Total", "Race", "Race", "Race", "Race",
                     "Race", "Race", "Race", "Gender", "Gender",
                     "Special Ed", "Special Ed", "ELL", "ELL")
```

Separate total from the remaining data:
```{r}
total <- table1[1,]
df <- table1[-1,]
```

We need to set up the category and characteristic columns as factors, which tells the plot to treat them as categories with a fixed number of options. The "characteristic" column is already factor, but R might order the factor alphabetically. In order to preserve the current order, we'll convert it to a character vector, and then back to a vector.
```{r}
df$category <- as.factor(df$category)
df$characteristic <- as.character(df$characteristic)
factor_order <- df$characteristic
df$characteristic <- factor(df$characteristic, levels=factor_order)

#create a column with the labels
#"White" is the topmost bar, so only it has the %
df$percent_label <- ifelse(df$characteristic == "White", 
                           paste0(round(df$percentageCTEparticipants, 0), "%"),
                           as.character(round(df$percentageCTEparticipants, 0)))
```

```{r}
#Most basic plot with no formatting
ggplot(df, aes(x=characteristic, y=percentageCTEparticipants, 
               fill=category)) + geom_bar(stat = "identity")

#stat = "identity" tells ggplot to use the y value as the height of the bar.
#Its default is stat = "count", which plots the number of occurrences
#of the x value, like a histogram.

#If we want to create a grouped bar chart, then we use stat = "identity"
#and add a second option, position. position = "dodge" will put the
#bars side by side, position = "stack" will stack them, and 
#position = "fill" will make a percent stacked barplot.
```

```{r}
#Set up list of color values and value of reference line for total
color_values = c("Race" = "#2EB4E7", "Gender" = "#7EC255", 
                 "Special Ed" = "#E7BE2E", "ELL" = "#E7612E")
ref_line = total$percentageCTEparticipants[1]

#Start off the same as with no formatting
plt <- ggplot(df, aes(x=characteristic, y=percentageCTEparticipants, fill=category)) + 
  geom_bar(stat="identity") + 
  #flip the coordinates so bars are horizontal
  coord_flip() +
  #set limits of y axis
  ylim(0, 93) +
  #set axis labels and titles
  labs(x="", y="", 
       title = "Percentage of high school students who were CTE participants") +
  #remove the legend, the gray background, and center title
  theme(legend.position = "none", 
        panel.background = element_blank(),
        plot.title = element_text(hjust=2)) + 
  #remove axis ticks (from the horizontal axis)
  guides(x="none") + 
  #set the order of the bars
  scale_x_discrete(limits=rev(levels(df$characteristic))) +
  #add labels
  geom_text(aes(label=percent_label), hjust=1) +
  #add reference line
  geom_hline(yintercept = ref_line, linetype=2) +
  #manually set the fill colors
  scale_fill_manual(values=color_values)

plt

```

```{r}
#this is a plotly function that takes a ggplot object
ggplotly(plt)
```



